Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1) \note{\scriptsize Authors, please add/check middle initials and affiliations}

### 2014-04-21 WSSSPE1_report/WSSSPE1_summary_report.tex
Revises Career Paths section.

Career path issues also came up repeatedly, starting in the the first keynote (\S\ref{sec:keynote1}), where Phil Bourne used the term ``the Google Bus'' to describe the issue of staff in his lab in San Diego who had chosen to leave science and work for Google (but since they didn't want to physically move, they take the Google bus to the Google office each day.)

The lack of career paths also came up in the Developing and Supporting Software discussion (\S\ref{sec:devel}), as an issue that led to difficulties in established projects retaining as well as recruiting staff. The issue there was that for software to be sustainable, there need to be career paths for software developers.  There was also some discussion about how should software development be balanced with research, with the idea that faculty are hired based on their research qualifications, not on their software development qualification, and as such, specializing in software development was likely to not lead to a career in science at a university, though clearly, national laboratories are a different case.

In the Policy discussion (\S\ref{sec:policy}), career paths were not discussed explicitly, though the discussion about credit, citation, and impact focused on how those who create or contribute to software are recognized; this recognition (or the current lack thereof) is tied to career paths (or their lack) for software developers.

The Communities discussion (\S\ref{sec:community}) did not directly address career paths, though one of the papers~\cite{Loffler_WSSSPE} that was part of this section also makes a similar policy point, in a section about career paths. Specifically, it says ``The most severe problem for developers in most computational sciences currently is that while most of the work is done creating hopefully well-written, sustainable software, the academic success is often exclusively tied to the solution of the scientific problem the software was designed for. Tasks that from a software engineering standpoint are essential, e.g., high usability, well-written and updated documentation, or porting infrastructure to new platforms, are not rewarded within this system.''

In terms of possible solutions, the Developing and Supporting Software discussion said that things are beginning to improve, for example, with incentives for software development in the form of altmetrics, tenure committee consideration, and NSF's change in its biosketch requirement: asking for proposer's ``products'' rather than their ``publications,'' as had been done previously.
However, it was also noted it can be very difficult to measure where people are using small bits of your code.  And the possible career paths for a specialist software developer working as part of a scientific research group in a university are certainly not yet clear.

>>>>

Career path issues also came up repeatedly, starting in the first
keynote (\S\ref{sec:keynote1}), where Phil Bourne used the term ``the
Google Bus'' to describe the issue of well-trained software development staff in
academic labs choosing to leave science and to work instead for technology
firms, especially Google, which happens in large enough numbers that
Google operates a bus every day to its nearest offices (and hence
staff who leaves academia in this way does not even have  to physically move).

The career path issue emerged repeatedly across panels because for
scientific software to be(come) sustainable, projects trying to create
sustainable software need to be able to recruit and retain software
developers trained in the various requisite software engineering
facets.  However, a career path in research means faculty at most
universitities, and as was noted repeatedly in discussions, faculty
are hired based on their scientific qualifications, not on their
software development skills or track record. Consequently, developing
special software development skills is unlikely to further a career in
science at a university, although national laboratories were
acknowledged as a different case. Loffler et
al.~\cite{Loffler_WSSSPE}, one of the papers in the Communities panel
(\S\ref{sec:community}), brought the problem to the point:
\begin{quote}
  ``The most severe problem for developers in most computational
  sciences currently is that while most of the work is done creating
  hopefully well-written, sustainable software, the academic success
  is often exclusively tied to the solution of the scientific problem
  the software was designed for. Tasks that from a software
  engineering standpoint are essential, e.g., high usability,
  well-written and updated documentation, or porting infrastructure to
  new platforms, are not rewarded within this system.'' ~\cite{Loffler_WSSSPE}
\end{quote}

Clearly, improving the recognition of software engineering work is
connected to addressing the career path problem. As was noted in the
Developing and Supporting Software (\S\ref{sec:devel}) and the Policy
(\S\ref{sec:policy}) panel discussions, there are encouraging signs of
improvement, including some altmetric services (such as Impactstory,
\url{http://impactstory.org}) collecting metrics for software source
code, and the fact that NSF now asks to list ``products'' rather than
their ``publications'' in an investigator's biosketch or results from
prior NSF support. However, how software, let alone parts of software
are reused by others can be very difficult to measure, and better
recognizing software products for principle investigators by itself
does not create career paths for specialist software developers
working as part of a university research group. Huang and
Lapp~\cite{Huang_WSSSPE}, a contribution to the Policy panel, offer
one possible solution in which a software engineering center of
excellence offers a career path for a correspondingly trained
workforce, and increased recognition of the resulting more sustainable
software would in a virtuous cycle heighten the value of the center's
services.
